---
title: "Capstone Project CKME 136"
author: 'Student: Walter Cios; Student ID 500791945'
output:
  html_document: default
  word_document: default
---


# Introduction

...

# Step 1: Data Cleaning and Preparation

### Load the package to read the raw data from Excel; load also ggplot2 for graphs; load dplyr for applying transformations on the dataframes; load package GGally to be used for plotting

```{r message = FALSE, warning = FALSE}
library(readxl)
library(ggplot2)
library(dplyr)
library(GGally)
```

### Loading the data 

After examining the original excel file, I noticed that only the first sheet had data, so I will only read
that from the file

```{r}
raw_data <- read_excel("./Data/ENB2012_data.xlsx", sheet = 1)
```

### Display structure of "raw_data"
I reviewed that all 768 observations are loaded

```{r}
str(raw_data)
```
### Check for NA in the whole dataset and found none 
....

```{r}
sum(sapply(raw_data, is.na))
```

As an alternative we could also have done an NA check per columns and as expected found none 

```{r}
sapply(raw_data, function(x) { sum(is.na(x)) })
```
### Change column name to real variable names 

....
```{r}
data <- raw_data
names(data) <- c("Relative Compactness", "Surface Area", "Wall Area", "Roof Area", "Overall Height",
                 "Orientation", "Glazing Area", "Glazing Area Distribution", "Heating Load", "Cooling Load")
```

## Converting categorical variables to factors

As per the dataset description the variable "Orientation" has only four values N,W,S,E 
The variable "Glazing Area Distribution"" has also the above and in addition "0" and "1"


```{r}
toCategorical <- Vectorize(function(column) {
  switch(as.character(column),
                      "0" = "No Glazing Area",
                      "1" = "Uniform",
                      "2" = "North",
                      "3" = "East",
                      "4" = "South",
                      "5" = "West",
                      NA)
})

data$`Glazing Area Distribution` <- as.factor(toCategorical(data$`Glazing Area Distribution`))
data$Orientation <- as.factor(toCategorical(data$Orientation))

```

## Review again structure, this time for "data"

```{r}
summary(data)
```

From the summmary we can see that Orientation and Glazing Area Distribution, are qualitative-categorical variables, the other variables are quantitaive-continuous


# Step 2: Exploratory Data Analysis 

Split up categorical variables from the continuous ones, into a new dataframe "data_numeric" 

```{r}
data_numeric <- select(data, -Orientation, -`Glazing Area Distribution`)
```

## Paired Scatterplots

I looked for a better display of plot and found the package "GGally" that is being used for scatterplots between all pairs of variables 

```{r fig.height=10, fig.width=10}
ggpairs(data_numeric)
```

# Spep 2 to be further detailed in the future anlaysis process 
